### 2020-11-13 latex/acrobot.tex
First editing pass of oscillation analysis

\section{Dynamics of the Acrobot}
Suppose we are given an acrobot as in Figure \ref{fig:acrobot-model} modelling a
gymnast hanging on a horizontal bar, where the ``torso" has moment of
inertia \(J_u\) and the ``leg" has moment of inertia \(J_a\) (each with respect
to their own center of mass).
Let \(q_u \in \mathbb{S}^1\) be the shoulder angle and \(q_a \in \mathbb{S}^1\) 
be the hip angle, where only \(q_a\) is actuated. 
Collecting them together provides the configuration
\(q = (q_u,q_a) \in \mathbb{S}^1 \times \mathbb{S}^1\). 
The acrobot has inertia matrix \(D\), potential function \(P\) (with respect to
the horizontal bar), and input matrix \(B\)  given as
follows \cite{xingbo_thesis}:
\begin{align}\label{eqn:general-acrobot-inertia}
    D(q) &= \begin{bmatrix}
      m_al_u^2 + 2m_a\cos(q_a)l_u l_{c_a} + m_al_{c_a}^2 + m_ul_{c_u}^2 + J_u + J_a &
      m_al_{c_a}^2 + m_al_ul_{c_a}\cos(q_a) + J_a \\
      m_al_{c_a}^2 + m_al_ul_{c_a}\cos(q_a) + J_a &
      m_al_{c_a}^2 + J_a
    \end{bmatrix} 
    , \\
    \label{eqn:general-acrobot-potential}
    P(q) &= g\left(m_al_{c_a}(1 - \cos(q_u+q_a)) + 
        (m_al_u + m_ul_{c_u})(1-\cos(q_u))\right) 
    , \\
    B(q) &= \begin{bmatrix} 0 \\ 1 \end{bmatrix}
    .
\end{align}

@@

When \(I = 0\), the constrained acrobot behaves like a single
pendulum with masses at a distance \(l\) and \(2l\) from the pivot 
(Figure \ref{fig:acrobot-I0}) whose energy
\begin{equation}\label{eqn:acrobot-nominal-E}
    E(q_u,p_u) = \frac{p_u^2}{10ml^2} + 3(1 - \cos(q_u))
    ,
\end{equation}
is conserved.
Level sets of \(E\) are ellipses when \(E < E(\pi,0)\), which we call
``oscillations"; and they are open curves when \(E > E(\pi,0)\), which we call
``rotations". 
Examples of these can be seen in Figure \ref{fig:pendulum-level-sets}.

@@

Drawing inspiration from \citet{dynamic_vhcs_stabilize_closed_orbits}, we want
to find a change of coordinates \((q_u,p_u) \to (\alpha, \mu)\) with the
following properties:
\begin{itemize}
    \item All oscillations in \((\alpha,\mu)\) coordinates are represented by
        \(\mu = \text{const}\).
    \item \(\alpha\) is a pseudo-angle living in \(\mathbb{S}^1\) which is
        always increasing.
\end{itemize}
We also want to find a second set of coordinates where the same properties hold for
rotations.
Once we have these coordinates in hand, we can use perturbation theory to prove
that \(\mu\) increases on average. 
If this is true, the acrobot must be gaining energy since the momentum is
increasing over time.

\subsection*{Background on Perturbation Theory}
Perturbation theory allows one to understand how solutions of nonlinear systems
behave by studying a nominal system which can be analyzed analytically.
The solutions of nonlinear systems are approximated by a Taylor-like expansion
around this nominal system.

\citet{khalil_nonlinear} considers a system of the form
\begin{equation}\label{eqn:khalil-setup}
    \begin{cases}
        \dot{x} = f(t,x,\epsilon), \\
        x(t_0) = \eta(\epsilon),
    \end{cases}
\end{equation}
where 
\(f : [t_0,t_1] \times D \times [-\epsilon_0,\epsilon_0] \rightarrow \R^n\) is
sufficiently smooth.
In the context of the acrobot, \(x\) is the phase in whichever coordinates make
our analysis convenient;
\(\epsilon\) is our control parameter \(I\);
the function \(f\) is the constrained dynamics in \(x\)-coordinates;
and \(\eta(\epsilon) \equiv \eta_0\) is a constant initial condition.
To keep a notational reminder that we are applying perturbation theory to
the acrobot, we will henceforth use \(I\) in place of \(\epsilon\).

Letting \(I = 0\) we get the nominal system
\begin{equation}\label{eqn:khalil-perturbation-nominal}
    \begin{cases}
        \dot{x}_0 = f(t,x,0) ,\\
        x_0(t_0) = \eta_0 ,
    \end{cases}
\end{equation}
which we need to solve for the explicit solution \(x_0(t,\eta_0)\) on \([t_0,t_1]\).
One can use this to compute the solution \(x_1(t,\eta_0)\) of
\begin{equation}\label{eqn:khalil-perturbation-firstorder}
    \begin{cases}
        \dot{x}_1 = \pdiff{f}{x}(t,x_0(t,\eta_0),0)x_1 + \pdiff{f}{I}(t,x_0(t,\eta_0),0)
        , \\
        x_1(t_0,\eta_0) = 0
        .
    \end{cases}
\end{equation}
Finally, one can compute the first-order approximation
\(x(t,\eta_0) = x_0(t,\eta_0) + I x_1(t,\eta_0) + R(t,I)\), where the remainder
term \(R(t,I)\) is \(O(I^2)\);
that is,
\[
    \lim \limits_{I \to 0} \frac{R(t,I)}{I} = 0
    .
\]
We now paraphrase Khalil's Theorem 10.1 \cite{khalil_nonlinear} on the
accuracy of perturbation analysis.
\begin{thm}
    Suppose \(f : [t_0,t_1] \times D \times [-I_0,I_0] \rightarrow \R^2\) is
    \(C^1\), and that the nominal problem ~\eqref{eqn:khalil-perturbation-nominal} has a
    unique solution \(x_0(t,\eta_0) \in D\) on \([t_0,t_1]\).
    Then there exists \(I^\star > 0\) such that for all \(|I| < I^\star\), the
    solution \(x(t,I,\eta_0)\) to ~\eqref{eqn:khalil-setup} satisfies
    \[
        \norm{x(t,I,\eta_0) - \left(x_0(t,\eta_0) + I x_1(t,\eta_))\right)} \leq k |I^2|
    \]
    for some \(k > 0\).
\end{thm}

The above Theorem tells us that we can approximate the solution of the nonlinear
system by the Taylor approximation 
\(x(t,\eta_0) \approx x_0(t,\eta_0) + I x_1(t,\eta_0)\).
In other words, when \(I\) is small enough solutions of the nonlinear system and
the Taylor approximation are the same up to order \(I^2\) along compact time
intervals.
Since we know that the acrobot behaves like a pendulum at \(I = 0\), we can
use this theory to more easily study the behaviour of orbits and prove energy
injection properties of our VNHC.

\subsection*{Perturbation Analysis for Oscillations}
When \(I = 0\), the acrobot becomes a nominal pendulum that oscillates whenever the
nominal energy ~\eqref{eqn:acrobot-nominal-E} is less than \(E(\pi,0)\).
Hence, the domain of oscillations on \(\Gamma\) is given by
\[
    D := \left\{ (q_u,p_u) \in \Gamma | E(q_u,p_u) < E(\pi,0)\right\}
    .
\]
We wish to find a transformation \(T(q_u,p_u)\) on \(D\) into
\((\alpha,\mu)\)-coordinates where \(\alpha \in \mathbb{S}^1\) is a pseudo-angle
and \(\mu > 0\) is a pseudo-radius.
Since the energy of oscillation can be determined by the orbit's intersection
with the \(q_u\)-axis and \(q_u \in ]-\pi,\pi[\) on \(D\), 
we set \(\mu \in ]0,\pi[\) as in Figure \ref{fig:mu-intersection}.
The transformation we want is therefore a diffeomorphism of the form
\begin{align*}
    T : D\backslash \{(0,0)\} &\rightarrow \mathbb{S}^1 \times \, ]0,\pi], \\
    (q_u, p_u) &\mapsto (\alpha,\mu)
    .
\end{align*}

\begin{figure}
    \centering
    % TODO: Add a figure with one oscillation mark at a distance mu from
    % the origin to the intersection on the q-axis. We also draw the boundary of
    % D and shade the interior
    %\includestandalone[]{images/mu_intersection}
    \caption{The domain \(D\) where a pendulum oscillates.
    The pseudo-radius coordinate \(\mu\) corresponds to the
    intersection of an orbit of oscillation with the \(q_u\)-axis.}
    \label{fig:mu-intersection}
\end{figure}

The energy level set corresponding to the intersection \((q_u,p_u) = (\mu,0)\)
is 
\[
    \left\{(q_u,p_u) \in \Gamma \mid E(q_u,p_u) = 3mgl(1- \cos(\mu))\right\}
    ,
\]
which gives the relationship
\begin{equation}\label{eqn:oscillation-pu2}
    p_u^2 = 30m^2gl^3(\cos(q_u) - \cos(\mu))
    .
\end{equation}
On this level set, \(q_u\) ranges between \([-\mu,\mu]\) and can be uniquely
parameterized by \(q_u = \mu \cos(\alpha)\), where \(\alpha\) is our
pseudo-angle.
Substituting this into ~\eqref{eqn:oscillation-pu2}, we get
\[
    p_u^2 = 30m^2gl^3(\cos(\mu \cos(\alpha)) - \cos(\mu))
    .
\]
We want to find \(p_u\) as a function of \((\alpha,\mu)\); noting that we can
determine the sign of \(p_u\) from the sign of \(\sin(\alpha)\), we get the
(clockwise) parameterization
\begin{equation}\label{eqn:oscillation-pu}
    p_u = -\sign{\sin(\alpha)} \sqrt{30m^2gl^3 \left(\cos(\mu c_\alpha) - c_\mu\right)}
    ,
\end{equation}
which is, in fact, smooth.

We have thus found a transformation \(T\inv(\alpha,\mu) = (q_u,p_u)\).
We need the inverse of this map to get our diffeomorphism \(T(q_u,p_u)\).
Notice from ~\eqref{eqn:oscillation-pu2} that
\[
    \cos(\mu) = -\frac{p_u^2}{30m^2gl^3} + \cos(q_u) =: C_\mu(q_u,p_u)
    .
\]
Since \(\mu \in ]0,\pi]\), we can uniquely express \(\mu\) by
\begin{equation}\label{eqn:mu-qupu}
    \mu = \arccos\left(C_\mu(q_u,p_u)\right)
    .
\end{equation}
Next we need to find \(\alpha\). 
Recall that 
\[
    \cos(\alpha) = \frac{q_u}{\mu}
    ,
\] 
which means 
\[
    \sin(\alpha) = \pm \sqrt{1 - \frac{q_u^2}{\mu^2}}
    .
\]
Notice from ~\eqref{eqn:oscillation-pu} that 
\(\sign{\sin(\alpha)} = -\sign{p_u}\).
We use these facts to deduce that
\begin{equation}\label{eqn:alpha-qupu}
    \alpha = \left.
        \arctan_2\left( -\sign{p_u}\sqrt{1 - \frac{q_u^2}{\mu^2}}, \frac{q_u}{\mu}\right)
        \right|_{\mu = \arccos(C_\mu(q_u,p_u))}
    ,
\end{equation}
and thus have our transformation \(T(q_u,p_u)\) into
\((\alpha,\mu)\)-coordinates on \(D\).

The acrobot's constrained dynamics in
\((\alpha,\mu)\)-coordinates can be computed by evaluating
\begin{align*}
    \dot{\mu} &= \left.\pdiff{\mu(q_u,p_u)}{q}_u\dot{q_u} +
         \pdiff{\mu(q_u,p_u)}{p_u}\dot{p}_u
         \right|_{(q_u,p_u) = T\inv(\mu,\alpha)}
    , \\
    \dot{\alpha} &= \left.\pdiff{\alpha(q_u,p_u)}{q_u}\dot{q}_u + 
        \pdiff{\alpha(q_u,p_u)}{p_u}\dot{p}_u
        \right|_{(q_u,p_u) = T\inv(\mu,\alpha)}
    .
\end{align*}
with \((\dot{q}_u,\dot{p}_u)\) the given by 
 ~\eqref{eqn:acrobot-constrained-dynamics}.

These dynamics are much too large to write out, so we simply denote them by
\begin{align}\label{eqn:acrobot-mu-dot}
    \dot{\mu} &= f_\mu(\alpha,\mu,I)
    ,\\
    \label{eqn:acrobot-alpha-dot}
    \dot{\alpha} &= f_\alpha(\alpha,\mu,I)
    .
\end{align}
The nominal system \((\alpha_n,\mu_n)\) comes from setting \(I = 0\).
MATLAB's symbolic toolbox evaluates the nominal dynamics as
\begin{align}\label{eqn:acrobot-mu-dot-nom}
    \dot{\mu}_n &= 0
    , \\
    \label{eqn:acrobot-alpha-dot-nom}
    \dot{\alpha}_n &= \sqrt{\frac{6g}{5l}} 
        \sqrt{\frac{\cos(\mu_n\cos(\alpha_n)) - \cos(\mu_n)}
            {\mu_n^2 \sin(\alpha_n)^2}}
    .
\end{align}
One can confirm that \(\dot{\alpha}_n > 0\) for every \(\mu_n \in ]0,\pi[\) -- this
is expected, since the nominal system is a single pendulum.
By continuity of ~\eqref{eqn:acrobot-mu-dot} with respect to \(I\), 
\(\dot{\alpha}\) remains positive for \(I\) small enough on the domain \(D\).

Suppose at time \(t = 0\) we have initialized the system at 
\((q_u,p_u) = (\mu_0,0)\), which corresponds to \((\alpha,\mu) = (0,\mu_0)\).
Since \(\alpha\) is continuously increasing, we know that the orbit will do one
of two things: it will either leave \(D\) so that the acrobot is now rotating,
or it will remain in \(D\) and hit the \(q_u\)-axis at some point 
\((\alpha,\mu) = (\pi,m)\) as is illustrated in Figure
\ref{fig:acrobot-possible-orbits}.
From there, the acrobot could escape \(D\) along the second half of the orbit or
return to some point \((\alpha,\mu) = (0,\tilde{m})\).
If \(\tilde{m} < \mu_0\) then the acrobot has lost energy; 
if \(\tilde{m} > \mu_0\) the acrobot has gained energy; 
and if \(\tilde{m} = \mu_0\), the acrobot is on a closed orbit.
Which situation occurs might depend entirely on the choice of
\(\mu_0\).

\begin{figure}
    \centering 
    % TODO: Draw the domain D, an initial point (q0,p0) = (mu0,0), and show the
    % 4 clockwise half orbits: when the orbit exits D on the bottom left corner;
    % when the orbit hits the left q-axis at (-mu0-x,0); when it hits (-mu0,0);
    % and when it hits (-mu0+x,0) for some x > 0.
    %includestandalone[width=0.5\textwdith]{images/acrobot_possible_orbits}
    \caption{If the acrobot is initialized at \((q_u,p_u) = (\mu_0,0)\), then
        either the orbit escapes \(D\) into the rotation
        zone (purple), or the orbit hits the other half of the \(q_u\)-axis 
        with a higher (red), equal (black), or lower (blue) angle than when it
        started.}
    \label{fig:acrobot-possible-orbits}
\end{figure}

Since \(\dot{\alpha} > 0\), we could theoretically find some time scaling 
\(t = \tau(\alpha)\) and use \(\alpha\) as a new ``time" variable.
Setting \(\hat{\mu}(\alpha) := \mu(\tau(\alpha))\) allows us to ignore the
dynamics of \(\alpha\) and study the evolution of \(\mu\) as a 
function of \(\alpha\) rather than as a function of \(t\) (assuming the acrobot
does not leave \(D\)).
The dynamics of \(\hat{\mu}\) are
\[
    \diff{\hat{\mu}}{\alpha} = 
    \left.\diff{\mu}{t} \diff{t}{\alpha} \right|_{t =\tau(\alpha)}
    ,
\] 
yielding the ``time"-varying scalar ODE
\begin{equation}\label{eqn:muhat-dot}
    \begin{cases}
        \diff{\hat{\mu}}{\alpha} 
        = \frac{f_\mu(\tau(\alpha),\mu,I)}{f_\alpha(\tau(\alpha),\mu,I)}
        =: g(\alpha,\mu,I)
        , \\
        \hat{\mu}(0) = \mu_0
        .
    \end{cases}
\end{equation}

In the remainder of this section we will find a first-order
approximation of \(\hat{\mu}\). 
We will show the dynamics of this approximation are \(\pi\)-periodic in
\(\alpha\), so that we only need to study whether \(\hat{\mu}\) increases or
decreases along a half-orbit \(\alpha \in [0,\pi]\).
To complete the proof of energy gain for oscillations, we will demonstrate that
\(\hat{\mu}(\pi) > \hat{\mu}(0)\).
This will imply that the the radius of orbits increases
until all orbits within \(D\) eventually leave \(D\) in finite time.

In the spirit of perturbation analysis, we expand our solution of
~\eqref{eqn:muhat-dot} into
\begin{equation}\label{eqn:acrobot-muhat-approx}
    \hat{\mu}(\alpha) = \hat{\mu}_0(\alpha) + I \hat{\mu}_1(\alpha)
    + R(\alpha,I)
    ,
\end{equation}
where \(R(\alpha,I)\) is \(O(I^2)\).
From ~\eqref{eqn:khalil-perturbation-nominal} we know that 
\(\hat{\mu}_0\) is the solution to the nominal system
\[
\begin{cases}
    \dot{\hat{\mu}}_0 = g(\alpha,\mu,0)
    , \\
    \hat{\mu}_0(0) = \mu_0
    ,
\end{cases}
\]
Since \(g(\alpha,\mu,0) = 0\) by 
~\eqref{eqn:acrobot-mu-dot-nom})-(\ref{eqn:acrobot-alpha-dot-nom}, we find that
\(\hat{\mu}_0(\alpha) = \mu_0\) for all \(\alpha\).

>>>>

\section{Dynamics of the Acrobot}
Suppose we are given an acrobot as in Figure \ref{fig:acrobot-model} modelling a
gymnast hanging on a horizontal bar, where the ``torso" has moment of
inertia \(J_u\) and the ``leg" has moment of inertia \(J_a\) (each with respect
to their own center of mass).
Let \(q_u \in \mathbb{S}^1\) be the shoulder angle and \(q_a \in \mathbb{S}^1\) 
be the hip angle, where only \(q_a\) is actuated. 
Collecting them together provides the configuration
\(q = (q_u,q_a) \in \mathbb{S}^1 \times \mathbb{S}^1\). 
The acrobot has inertia matrix \(mathcal{O}\), potential function \(P\) (with respect to
the horizontal bar), and input matrix \(B\)  given as
follows \cite{xingbo_thesis}:
\begin{align}\label{eqn:general-acrobot-inertia}
    D(q) &= \begin{bmatrix}
      m_al_u^2 + 2m_a\cos(q_a)l_u l_{c_a} + m_al_{c_a}^2 + m_ul_{c_u}^2 + J_u + J_a &
      m_al_{c_a}^2 + m_al_ul_{c_a}\cos(q_a) + J_a \\
      m_al_{c_a}^2 + m_al_ul_{c_a}\cos(q_a) + J_a &
      m_al_{c_a}^2 + J_a
    \end{bmatrix} 
    , \\
    \label{eqn:general-acrobot-potential}
    P(q) &= g\left(m_al_{c_a}(1 - \cos(q_u+q_a)) + 
        (m_al_u + m_ul_{c_u})(1-\cos(q_u))\right) 
    , \\
    B(q) &= \begin{bmatrix} 0 \\ 1 \end{bmatrix}
    .
\end{align}

@@

When \(I = 0\), the constrained acrobot behaves like a single
pendulum with masses at a distance \(l\) and \(2l\) from the pivot 
(Figure \ref{fig:acrobot-I0}) whose energy
\begin{equation}\label{eqn:acrobot-nominal-E}
    E(q_u,p_u) = \frac{p_u^2}{10ml^2} + 3(1 - \cos(q_u))
    ,
\end{equation}
is conserved.
Level sets of \(E\) are ellipses when \(E(q_u,p_u) < E(\pi,0)\), which we call
``oscillations"; and they are open curves when \(E(q_u,p_u) > E(\pi,0)\), which we call
``rotations". 
Examples of these can be seen in Figure \ref{fig:pendulum-level-sets}.

@@

Using a method developed by \citet{dynamic_vhcs_stabilize_closed_orbits},
we can find a change of coordinates \((q_u,p_u) \to (\alpha, \mu)\) with the
following properties:
\begin{itemize}
    \item The energy of oscillation in \((\alpha,\mu)\) coordinates is 
        uniquely defined by \(\mu\), which remains constant along oscillations
        of the simple pendulum.
    \item \(\alpha\) is a pseudo-angle living in \(\mathbb{S}^1\) which is
        always increasing.
\end{itemize}
Once we have these coordinates in hand, we can use perturbation theory to prove
that \(\mu\) increases on average for the acrobot when \(I\) is small enough.
Likewise, we will find a second set of coordinates \((\beta, \rho)\) where
\(\beta \in \mathbb{S}^1\) is always increasing, \(\rho\) is constant along
rotations of the simple pendulum, and \(\rho\) increases for the acrobot when
\(I\) is small enough.
Putting these together, we will show the acrobot is gaining energy.

\subsection*{Background on Perturbation Theory}
Nonlinear systems like the acrobot are difficult (or impossible) to solve
analytically.
Perturbation theory allows one to understand the behaviour of the nonlinear system
by studying a simpler nominal system. 
Solutions of the nonlinear system can often be approximated by taking a Taylor
expansion around this nominal system.

\citet{khalil_nonlinear} considers a system of the form
\begin{equation}\label{eqn:khalil-setup}
    \begin{cases}
        \dot{x} = f(t,x,\epsilon), \\
        x(t_0) = \eta(\epsilon),
    \end{cases}
\end{equation}
where 
\(f : [t_0,t_1] \times D \times [-\epsilon_0,\epsilon_0] \rightarrow \R^n\) is
sufficiently smooth.
In the context of the acrobot, \(x\) is the set of coordinates which make
our analysis convenient;
\(\epsilon\) is our control parameter \(I\);
the function \(f\) is the constrained dynamics in \(x\)-coordinates;
and \(\eta(\epsilon) \equiv \eta_0\) is a constant initial condition.
To keep a notational reminder that we are applying perturbation theory to
the acrobot, we will henceforth use \(I\) in place of \(\epsilon\).

Setting \(I = 0\) we get the nominal system
\begin{equation}\label{eqn:khalil-perturbation-nominal}
    \begin{cases}
        \dot{x}_0 = f(t,x,0) ,\\
        x_0(t_0) = \eta_0 ,
    \end{cases}
\end{equation}
which we need to solve for the explicit solution \(x_0(t,\eta_0)\) on \([t_0,t_1]\).
One can use this to compute the solution \(x_1(t,\eta_0)\) of
\begin{equation}\label{eqn:khalil-perturbation-firstorder}
    \begin{cases}
        \dot{x}_1 = \pdiff{f}{x}(t,x_0(t,\eta_0),0)x_1 + \pdiff{f}{I}(t,x_0(t,\eta_0),0)
        , \\
        x_1(t_0,\eta_0) = 0
        .
    \end{cases}
\end{equation}
Finally, one can compute the first-order Taylor approximation
\(x(t,\eta_0) = x_0(t,\eta_0) + I x_1(t,\eta_0) + R(t,I)\), where the remainder
term \(R(t,I)\) is \(O(I^2)\), \ie,
\[
    \lim \limits_{I \to 0} \frac{R(t,I)}{I} = 0
    .
\]
We now paraphrase Khalil's Theorem 10.1 \cite{khalil_nonlinear} on the
accuracy of perturbation analysis.
\begin{thm}\label{thm:khalil-perturbation}
    Suppose \(f : [t_0,t_1] \times D \times [-I_0,I_0] \rightarrow \R^2\) is
    \(C^1\), and that the nominal problem ~\eqref{eqn:khalil-perturbation-nominal} has a
    unique solution \(x_0(t,\eta_0) \in D\) on \([t_0,t_1]\).
    Then there exists \(I^\star > 0\) such that for all \(I,\, |I| < I^\star\), the
    solution \(x(t,I,\eta_0)\) to ~\eqref{eqn:khalil-setup} satisfies
    \[
        \norm{x(t,I,\eta_0) - \left(x_0(t,\eta_0) + I x_1(t,\eta_0)\right)} \leq k |I^2|
    \]
    for some \(k > 0\).
\end{thm}

Theorem \ref{thm:khalil-perturbation} tells us that we can approximate the
solution of the nonlinear system by the Taylor approximation 
\(x(t,\eta_0) \approx x_0(t,\eta_0) + I x_1(t,\eta_0)\).
When \(I\) is small enough, solutions of the nonlinear system and
the Taylor approximation are the same up to order \(I^2\) along compact time
intervals.
Since we know that the acrobot behaves like a pendulum at \(I = 0\), we can
use this theory to prove energy injection properties on the acrobot by studying
the simple pendulum.

\subsection*{Perturbation Analysis for Oscillations}
As we have seen, the acrobot becomes a nominal pendulum that oscillates whenever the
nominal energy ~\eqref{eqn:acrobot-nominal-E} is less than \(E(\pi,0)\).
Hence, the domain of oscillations for the nominal pendulum is given by
\[
    \mathcal{O} := \left\{ (q_u,p_u) \in \Gamma \mid E(q_u,p_u) < E(\pi,0)\right\}
    ,
\]
where \(\Gamma = \mathbb{S}^1\times\R\) is the consraint manifold of the
acrobot.
We wish to find a transformation \(T(q_u,p_u)\) on \(\mathcal{O}\) into
\((\alpha,\mu)\)-coordinates where \(\alpha \in \mathbb{S}^1\) is a pseudo-angle
and \(\mu > 0\) is a pseudo-radius.
Since the energy of oscillation can be determined by the orbit's intersection
with the \(q_u\)-axis (as in Figure \ref{fig:mu-intersection}) 
and \(q_u \in ]-\pi,\pi[\) on \(\mathcal{O}\), we set \(\mu \in ]0,\pi[\).
The transformation we want is therefore a diffeomorphism of the form
\begin{align*}
    T : D\backslash \{(0,0)\} &\rightarrow \mathbb{S}^1 \times \, ]0,\pi[, \\
    (q_u, p_u) &\mapsto (\alpha,\mu)
    .
\end{align*}

\begin{figure}
    \centering
    % TODO: Add a figure with one oscillation mark at a distance mu from
    % the origin to the intersection on the q-axis. We also draw the boundary of
    % D and shade the interior
    %\includestandalone[]{images/mu_intersection}
    \caption{The domain \(\mathcal{O}\) where a pendulum oscillates.
    The pseudo-radius coordinate \(\mu\) corresponds to the
    intersection of an orbit of oscillation with the \(q_u\)-axis.}
    \label{fig:mu-intersection}
\end{figure}

The energy level set corresponding to the intersection \((q_u,p_u) = (\mu,0)\)
is 
\[
    \left\{(q_u,p_u) \in \Gamma \mid E(q_u,p_u) = 3mgl(1- \cos(\mu))\right\}
    ,
\]
which gives the relationship
\begin{equation}\label{eqn:oscillation-pu2}
    p_u^2 = 30m^2gl^3(\cos(q_u) - \cos(\mu))
    .
\end{equation}
On this level set, \(q_u\) ranges between \([-\mu,\mu]\) and can be uniquely
parameterized by \(q_u = \mu \cos(\alpha)\), where \(\alpha\) is our
desired pseudo-angle.
Substituting this into ~\eqref{eqn:oscillation-pu2}, we get
\[
    p_u^2 = 30m^2gl^3(\cos(\mu \cos(\alpha)) - \cos(\mu))
    .
\]
We want to find \(p_u\) as a function of \((\alpha,\mu)\); noting that we can
determine the sign of \(p_u\) from the sign of \(\sin(\alpha)\), we get the
(clockwise) parameterization
\begin{equation}\label{eqn:oscillation-pu}
    p_u = -\sign{\sin(\alpha)} \sqrt{30m^2gl^3 \left(\cos(\mu c_\alpha) - c_\mu\right)}
    ,
\end{equation}
which is smooth for all \(\mu \in ]0,\pi[\).

We have thus found a transformation \(T\inv(\alpha,\mu) = (q_u,p_u)\).
We need the inverse of this map to get our diffeomorphism \(T(q_u,p_u)\).
Notice from ~\eqref{eqn:oscillation-pu2} that
\[
    \cos(\mu) = -\frac{p_u^2}{30m^2gl^3} + \cos(q_u) =: C_\mu(q_u,p_u)
    .
\]
Since \(\mu \in ]0,\pi[\), we can uniquely express \(\mu\) by
\begin{equation}\label{eqn:mu-qupu}
    \mu = \arccos\left(C_\mu(q_u,p_u)\right)
    .
\end{equation}
Next we need to find \(\alpha\). 
Recall that 
\begin{equation}\label{eqn:acrobot-cosalpha-qu}
    \cos(\alpha) = \frac{q_u}{\mu}
    ,
\end{equation}
which means 
\begin{equation}\label{eqn:acrobot-sinalpha-qu}
    \sin(\alpha) = \pm \sqrt{1 - \frac{q_u^2}{\mu^2}}
    .
\end{equation}
Using ~\eqref{eqn:oscillation-pu}, we determine that
\(\sign{\sin(\alpha)} = -\sign{p_u}\).
Putting together ~\eqref{eqn:acrobot-cosalpha-qu} and
~\eqref{eqn:acrobot-sinalpha-qu}, we deduce that
\begin{equation}\label{eqn:alpha-qupu}
    \alpha = \left.
        \arctan_2\left( -\sign{p_u}\sqrt{1 - \frac{q_u^2}{\mu^2}}, \frac{q_u}{\mu}\right)
        \right|_{\mu = \arccos(C_\mu(q_u,p_u))}
    .
\end{equation}
Thus, ~\eqref{eqn:mu-qupu} and ~\eqref{eqn:alpha-qupu} define 
our transformation into \((\alpha,\mu)\)-coordinates on \(\mathcal{O}\).

The acrobot's constrained dynamics in
\((\alpha,\mu)\)-coordinates can be computed by evaluating
\begin{align*}
    \dot{\mu} &= \left.\pdiff{\mu(q_u,p_u)}{q}_u\dot{q_u} +
         \pdiff{\mu(q_u,p_u)}{p_u}\dot{p}_u
         \right|_{(q_u,p_u) = T\inv(\mu,\alpha)}
    , \\
    \dot{\alpha} &= \left.\pdiff{\alpha(q_u,p_u)}{q_u}\dot{q}_u + 
        \pdiff{\alpha(q_u,p_u)}{p_u}\dot{p}_u
        \right|_{(q_u,p_u) = T\inv(\mu,\alpha)}
    .
\end{align*}
with \((\dot{q}_u,\dot{p}_u)\) given by 
 ~\eqref{eqn:acrobot-constrained-dynamics}.

These dynamics are much too large to write out, so we simply denote them by
\begin{align}\label{eqn:acrobot-mu-dot}
    \dot{\mu} &= f_\mu(\alpha,\mu,I)
    ,\\
    \label{eqn:acrobot-alpha-dot}
    \dot{\alpha} &= f_\alpha(\alpha,\mu,I)
    .
\end{align}
The dynamics of the nominal pendulum arise from evaluating
\(f_\mu\) and \(f_\alpha\) at \(I = 0\).
MATLAB's symbolic toolbox evaluates the nominal dynamics as
\begin{align}\label{eqn:acrobot-mu-dot-nom}
    \dot{\mu}_n &= 0
    , \\
    \label{eqn:acrobot-alpha-dot-nom}
    \dot{\alpha}_n &= \sqrt{\frac{6g}{5l}} 
        \sqrt{\frac{\cos(\mu_n\cos(\alpha_n)) - \cos(\mu_n)}
            {\mu_n^2 \sin(\alpha_n)^2}}
    .
\end{align}
Note that we write the nominal coordinates as \((\alpha_n,\mu_n)\) to
distinguish them from the acrobot's coordinates.
One can confirm that \(\dot{\alpha}_n > 0\) for every \(\mu_n \in ]0,\pi[\).
This is expected of an oscillating single pendulum.
By continuity of ~\eqref{eqn:acrobot-mu-dot}, 
\(\dot{\alpha}\) remains positive on \(\mathcal{O}\) for \(I\) small enough.

Suppose at time \(t = 0\) we have initialized the acrobot at 
\((q_u,p_u) = (\mu_0,0)\), which corresponds to \((\alpha,\mu) = (0,\mu_0)\).
Since \(\alpha\) is continuously increasing, we know that the orbit will do one
of two things: it will either leave \(\mathcal{O}\) so that the acrobot is now rotating,
or it will remain in \(\mathcal{O}\) and hit the \(q_u\)-axis at some point 
\((\alpha,\mu) = (\pi,m)\) (as illustrated in Figure
\ref{fig:acrobot-possible-orbits}).
From there, the acrobot could escape \(\mathcal{O}\) along the second half of the orbit or
return to some point \((\alpha,\mu) = (0,\tilde{m})\).
If \(\tilde{m} < \mu_0\) then the acrobot has lost energy; 
if \(\tilde{m} > \mu_0\) the acrobot has gained energy; 
and if \(\tilde{m} = \mu_0\), the acrobot is on a closed orbit.
Which situation occurs might depend entirely on the choice of
\(\mu_0\).

\begin{figure}
    \centering 
    % TODO: Draw the domain D, an initial point (q0,p0) = (mu0,0), and show the
    % 4 clockwise half orbits: when the orbit exits D on the bottom left corner;
    % when the orbit hits the left q-axis at (-mu0-x,0); when it hits (-mu0,0);
    % and when it hits (-mu0+x,0) for some x > 0.
    %includestandalone[width=0.5\textwdith]{images/acrobot_possible_orbits}
    \caption{If the acrobot is initialized at \((q_u,p_u) = (\mu_0,0)\), then
        either the orbit escapes \(\mathcal{O}\) into the rotation
        zone (purple), or the orbit hits the other half of the \(q_u\)-axis 
        with a higher (red), equal (black), or lower (blue) angle than when it
        started.}
    \label{fig:acrobot-possible-orbits}
\end{figure}

Since \(\dot{\alpha} > 0\), we could theoretically find some time scaling 
\(t = \tau(\alpha)\) and use \(\alpha\) as a new ``time" variable.
Setting \(\hat{\mu}(\alpha) := \mu(\tau(\alpha))\) allows us to ignore the
dynamics of \(\alpha\) and study the evolution of \(\mu\) as a 
function of \(\alpha\) rather than as a function of \(t\) (assuming the acrobot
does not leave \(\mathcal{O}\)).
The dynamics of \(\hat{\mu}\) are
\[
    \diff{\hat{\mu}}{\alpha} = 
    \left.\diff{\mu}{t} \diff{t}{\alpha} \right|_{t =\tau(\alpha)}
    ,
\] 
yielding the ``time"-varying scalar ODE
\begin{equation}\label{eqn:muhat-dot}
    \begin{cases}
        \diff{\hat{\mu}}{\alpha} 
        = \frac{f_\mu(\tau(\alpha),\mu,I)}{f_\alpha(\tau(\alpha),\mu,I)}
        =: g(\alpha,\mu,I)
        , \\
        \hat{\mu}(0) = \mu_0
        .
    \end{cases}
\end{equation}
with solution \(\hat{\mu}(\alpha,\mu_0,I)\).

%Using perturbation theory, we will find a first-order
%approximation of \(\hat{\mu}\). 
%We will show the dynamics of this approximation are \(\pi\)-periodic in
%\(\alpha\), so that we only need to study whether \(\hat{\mu}\) increases or
%decreases along a half-orbit \(\alpha \in [0,\pi]\).
%To complete the proof of energy gain for oscillations, we will demonstrate that
%\(\hat{\mu}(\pi) > \hat{\mu}(0)\).
%This will imply that the the radius of orbits increases
%until all orbits within \(\mathcal{O}\) eventually leave \(\mathcal{O}\) in finite time.
